### 2015-05-10 swmpr_manu.Rnw
cover letter added

\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \gls{NERRS}.  This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.  \gls{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with \gls{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
\linenumbers
\glsaddall
\printglossaries

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acrlong{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \gls{SWMP} data.  \gls{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \gls{SWMP} \cite{SWMP14}.

@@

\section*{Summary}

@@

\subsection*{Sample data files}
\label{supp_zip}

\subsection*{\gls{SWMP} data as R binary files}

\subsection*{Metabolism scripts}
\label{met_scripts}

\subsection*{Source code for reproducing figures/tables}

\subsection*{Data to reproduce figures/tables}

@@

I acknowledge the significant work of \gls{NERRS} researchers and staff that has allowed access to high-quality monitoring data.  Thanks to Dwayne Porter and Melissa Ide from \gls{CDMO} for maintaining the online database.  Thanks to Marie Bundy and Nikki Dix for giving me the opportunity to share this package with the broader \gls{NERRS} community.  Thanks to Todd O'Brien for the inspiration for the online widgets. Thanks to Mike Murrell and Jim Hagy III for assistance with documentation and implementation of the ecosystem metabolism functions. The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.  The use of trade names or products does not constitute endorsement by the US Government.

@@

>>>>

\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \acrlong{NERRS}.  This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.  \gls{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with \gls{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
\linenumbers
\glsaddall
\printglossaries

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acrlong{SWMP} (\acrshort{SWMP}) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \gls{SWMP} data.  \gls{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \gls{SWMP} \cite{SWMP14}.

@@

\section*{Conclusions}

@@

@@

I acknowledge the significant work of \gls{NERRS} researchers and staff that has allowed access to high-quality monitoring data.  Thanks to Todd O'Brien for the inspiration for the online widgets. Thanks to Mike Murrell and Jim Hagy III for assistance with the ecosystem metabolism functions. Thanks to Jeffrey Hollister for providing useful comments on an earlier draft.  The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.  The use of trade names or products does not constitute endorsement by the US Government.

@@

\subsection*{S1 Sample data files}
\label{supp_zip}

\subsection*{S2 \gls{SWMP} data as R binary files}

\subsection*{S3 Metabolism scripts}
\label{met_scripts}

\subsection*{S4 Source code for reproducing figures/tables}

\subsection*{S5 Data to reproduce figures/tables}
